---
title: "ENVT5503 time series in R"
documentclass: article
classoption: a4paper
geometry: margin=0.787in
output: 
  bookdown::html_document2:
    fig_width: 6.4
    fig_height: 4.8
    number_sections: no
    self_contained: no
    df_print: default
    fig_caption: yes
    smart: no
  html_document: 
    fig_width: 6.4
    fig_height: 4.8
    number_sections: no
    self_contained: no
    df_print: default
    fig_caption: yes
    smart: no
  word_document: 
    toc_depth: 2
    fig_caption: yes
editor_options: 
  markdown: 
    wrap: 80
---

```{r page 1 header hide code, fig.height=1.5, fig.width=10, echo=FALSE, out.width="100%", fig.align='right', results='hold'}
library(png)
addImg <- function(obj, x = NULL, y = NULL, width = NULL, interpolate = TRUE){
  if(is.null(x) | is.null(y) | is.null(width)){stop("Must provide args 'x', 'y', and 'width'")}
  USR <- par()$usr ; PIN <- par()$pin ; DIM <- dim(obj) ; ARp <- DIM[1]/DIM[2]
  WIDi <- width/(USR[2]-USR[1])*PIN[1] ;   HEIi <- WIDi * ARp 
  HEIu <- HEIi/PIN[2]*(USR[4]-USR[3]) 
  rasterImage(image = obj, xleft = x-(width/2), xright = x+(width/2),
            ybottom = y-(HEIu/2), ytop = y+(HEIu/2), interpolate = interpolate)
}
logo <- readPNG("UWA logo_text_V_wsL.png")
par(mar = c(0,0,0,0))
layout(matrix(c(1,1,1,1,1,2),nrow = 1))

plot(c(0,1),c(0,1), axes=F, type="n",xaxt="n", yaxt="n",ann=F)
text(-0.025,0.9, pos = 4, cex = 2.6, font = 2, 
     labels="Data Analysis in R for Environmental Science")
text(-0.025,0.6, pos = 4, cex = 2.2, 
     labels="ENVT5503 Time Series")
text(-0.025,0.4, pos = 4, font = 3, cex = 1.4, col = "navy",
     labels="Learning time series analysis concepts using hourly soil temperature data")
text(1.04,0.1, pos = 2, font = 3, family = 'serif', cex = 1.5, col = "navy",
     labels="Andrew Rate, School of Agriculture and Environment")
plot(1,1, axes=F, type="n",xaxt="n", yaxt="n",ann=F)
addImg(logo, x = 1, y = 1, width = 0.6)
par(mar = c(3.5,3.5,0.5,0.5))
```

# Set up the R environment for time series analysis

We will need the additional functions in several **R** packages for specialized
time series and other functions in R...

```{r setup run, message=FALSE, warning=FALSE, include=TRUE, results='hide'}
# Load the packages we need 
library(zoo)      # for basic irregular time series functions
library(xts)      # we need the xts "eXtended Time Series" format for some functions
library(Kendall)  # for trend analysis with Mann-Kendall test
library(trend)    # for trend analysis using the Sen slope
library(forecast) # for time series forecasting with ARIMA and exponential smoothing 
library(tseries)  # for assessing stationarity using Augmented Dickey-Fuller test
library(lmtest)   # for Breusch-Pagan heteroscedasticity test etc.
library(car)      # for various commonly-used functions
library(ggplot2)  # alternative to base R plots

# (optional) make a better colour palette than the R default!
palette(c("black","red3","green3","blue2",
          "darkcyan","purple","sienna","gray50"))
par(mfrow = c(3,1), mar = c(4,4,1,1), mgp = c(1.7,0.3,0), tcl = 0.3, font.lab=2)
```

# Data input

Read the data into a data frame - this is how R often stores data - it's not the
format we need but we'll use it for comparison.

## Non-time series object for comparison

```{r read into dataframe}
soiltemp <- read.csv("soiltemp2.csv")
colnames(soiltemp) <- c("Date","temp")
```

## Do some checks of the data

```{r check dataframe}
summary(soiltemp) # simple summary of each column
str(soiltemp) # more detailed information about the R object ('str'=structure)
```

### Check with a plot

```{r plot-df, out.width="50%", fig.align='center', fig.cap="Plot of soil temperature time series data which is not (yet) formatted as a time series.", results='hold'}
plot(soiltemp$temp, type = "l", col = 6)
```

The horizontal axis in Figure \@ref(fig:plot-df) is just the row number of the
data frame, not a date or time. We really want the data in a different type of R
object! We use the [read.csv.zoo()]{style="font-family: monospace;
background-color: #e8e8e8;"} function in the [zoo]{style="font-family:
monospace; background-color: #e8e8e8;"} package to read the data in as a time
series. This gives a data object of class 'zoo' which is much more flexible than
the default time series object in R.

```{r read and summary csv to zoo}
soiltemp_T15_zoo <- read.csv.zoo("soiltemp2.csv",
                               format = "%Y-%m-%d %H:%M:%S", 
                               tz = "Australia/Perth", 
                               index.column=1,
                               header = TRUE)
# do some quick checks of the new R object:
summary(soiltemp_T15_zoo) 
str(soiltemp_T15_zoo) # POSIXct in the output is a date-time format
```

It's usually useful to check our data with a plot (Figure \@ref(fig:plot-zoo))

```{r plot-zoo, out.width="50%", fig.align='center', fig.cap="Plot of soil temperature time series data which is formatted as a zoo time series object.", results='hold'}
plot(soiltemp_T15_zoo, col = 2) 
```

Sometimes we need to use another time series data format,
[xts]{style="font-family: monospace; background-color: #e8e8e8;"}
(e[X]{.underline}tended [T]{.underline}ime [S]{.underline}eries) which allows us
to use more functions...

### Make an xts object from our zoo object and check:

```{r make xts from zoo and check}
soiltemp_T15_xts <- as.xts(soiltemp_T15_zoo)
str(soiltemp_T15_xts) # just to check
```

```{r plot-xts, out.width="60%", fig.align='center', fig.cap="Plot of soil temperature time series data which is formatted as an xts time series object.", results='hold'}
plot(soiltemp_T15_xts, col = 6, ylab = "Soil temperature (\u00B0C)") # just to check
```

The plot in Figure \@ref(fig:plot-xts) shows that plotting an 
[xts]{style="font-family: monospace; background-color: #e8e8e8;"} time series 
object gives a somewhat more detailed plot than for a 
[zoo]{style="font-family: monospace; background-color: #e8e8e8;"} formatted 
time series object (Figure \@ref(fig:plot-zoo)).

# Exploratory data analysis of time series

We first examine a plot of the data in our time series object, both the raw
data and using common transformations (Figure \@ref(fig:plot-transf)). We
compare with transformed time series data - we may need to do this to meet later
modelling assumptions.

First we change the default plotting parameters using
[par(...)]{style="font-family: monospace; background-color: #e8e8e8;"}, then
plot the object and some transformed versions (with custom axis labels).

```{r plot-transf, fig.height=7, fig.width=7, out.width="65%", fig.align='center', fig.cap="Plots of soil temperature time series data as untransformed, log10-transformed, and power transformed values.", results='hold'}
par(mfrow = c(3,1), mar = c(4,4,1,1), mgp = c(1.7,0.3,0), tcl = 0.3, font.lab=2)

plot(soiltemp_T15_zoo, ylab = "Temperature (\u00B0C)",
     xlab = "Date", col = 7, lwd = 2, cex.lab = 1.4)

plot(log10(soiltemp_T15_zoo), 
     ylab = expression(bold(paste(log[10],"(Temperature, \u00B0C)"))),
     xlab = "Date", col = 6, lwd = 2, cex.lab = 1.4)

pt0 <- powerTransform(coredata(soiltemp_T15_zoo))
if(pt0$lambda<0) {
  plot(
    -1 * (soiltemp_T15_zoo ^ pt0$lambda),
    ylab = "power-transf. (Temp., \u00B0C)",
    xlab = "Date",
    col = 4, lwd = 2, cex.lab = 1.4
  )
} else {
  plot((soiltemp_T15_zoo ^ pt0$lambda),
       ylab = "power-transf. (Temp., \u00B0C)",
       xlab = "Date",
       col = 4, lwd = 2, cex.lab = 1.4
  )
}
```

```{r remove-pt0, include=FALSE}
rm(pt0)
```

Which of these plots looks like it might be homoscedastic, *i.e*., have constant
variance regardless of time?

The R Cookbook suggests Box-Cox (power) transforming the variable to "stabilize
the variance" (*i.e*. reduce heteroscedasticity). This does work, but can be
better to use log<sub>10</sub>[conc], as the values are easier to interpret.

If we think a transformation is needed, then run something like the code below
(which does a square root transformation, *i.e*. variable<sup>0.5</sup>). This
is just an example -- we may, for example, decide that the
log<sub>10</sub>-transformation is more suitable.

```{r transform if needed not run, eval=FALSE, echo=TRUE}
soiltemp_T15_zoo <- soiltemp_T15_zoo^0.5 
# don't forget the xts version either!
soiltemp_T15_xts <- as.xts(soiltemp_T15_zoo)
```

# Assessing if a time series variable is stationary

A **stationary** variable's mean and variance are not dependent on time. In
other words, for a stationary series, the mean and variance at any time are
representative of the whole series.

If there is a trend for the value of the variable to increase or decrease, or if
there are periodic fluctuations, we don't have a stationary time series.

Many useful statistical analyses and models for time series models need a
stationary time series as input, or a time series that can be made stationary
with transformations or differencing.

## Testing for stationarity

```{r adf tests}
# we need the package 'tseries' for the Augmented Dickey–Fuller (adf) Test
d0 <- adf.test(soiltemp_T15_zoo); print(d0)
d1 <- adf.test(diff(soiltemp_T15_zoo,1)); print(d1)
d2 <- adf.test(diff(diff(soiltemp_T15_zoo,1),1)); print(d2)
```

## Plot differencing for stationarity

The plots in Figure \@ref(fig:plot-diff) below show that constant mean and 
variance (*i.e*. stationarity) is achieved after a single differencing step. 
The [adf.test]{style="font-family: monospace; background-color: #e8e8e8;"} 
p-value suggests that the non-differenced series is already stationary, but the 
first sub-plot in Figure \@ref(fig:plot-diff) does not support this.

```{r plot-diff, fig.height=8, fig.width=8, out.width="65%", fig.align='center', fig.cap="Plots of soil temperature time series data and its first and second differences, with each sub-plot showing the smoothed (Loess) and linear trends.", results='hold'}
par(mfrow = c(3,1), mar = c(0,4,0,1), oma = c(4,0,1,0), cex.lab = 1.4,
    mgp = c(2.5,0.7,0), font.lab = 2)
plot(soiltemp_T15_zoo, ylab = "Raw data, no differencing",
     xlab="", xaxt="n", col = 8)
lines(loess.smooth(index(soiltemp_T15_zoo),coredata(soiltemp_T15_zoo)), 
      col = 4, lwd = 2)
abline(lm(coredata(soiltemp_T15_zoo) ~ index(soiltemp_T15_zoo)), col = 2)
legend("topright", legend = c("soiltemp_T15_Data", "Loess smoothing","Linear model"),
       cex = 1.8, col = c(1,4,2), lwd = c(1,2,1), bty = "n")
mtext(paste("adf.test p value =",signif(d0$p.value,3)), 
      side = 1, line = -1.2, adj = 0.05)
plot(diff(soiltemp_T15_zoo,1),
     ylab = "First differencing",
     xlab="", xaxt="n", col = 8)
abline(h = 0, col = "grey", lty = 2)
lines(loess.smooth(index(diff(soiltemp_T15_zoo,1)),coredata(diff(soiltemp_T15_zoo,1))), 
      col = 4, lwd = 2)
abline(lm(coredata(diff(soiltemp_T15_zoo,1)) ~
            index(diff(soiltemp_T15_zoo,1))), col = 2)
mtext(paste("adf.test p value =",signif(d1$p.value,3)), 
      side = 1, line = -1.2, adj = 0.05)
plot(diff(diff(soiltemp_T15_zoo,1),1),
     ylab = "Second differencing",
     xlab="Date", col = 8)
mtext("Date",side = 1, line = 2.2, font = 2)
abline(h = 0, col = "grey", lty = 2)
lines(loess.smooth(index(diff(diff(soiltemp_T15_zoo,1),1)),
                   coredata(diff(diff(soiltemp_T15_zoo,1),1))), 
      col = 4, lwd = 2)
abline(lm(coredata(diff(diff(soiltemp_T15_zoo,1))) ~ 
            index(diff(diff(soiltemp_T15_zoo,1)))), col = 2)
mtext(paste("adf.test p value =",signif(d2$p.value,3)), 
      side = 1, line = -1.2, adj = 0.05)
```

```{r reset mfrow 1, message=FALSE, warning=FALSE, include=FALSE}
par(mfrow = c(1,1), mar = c(4,4,1,1), oma = c(0,0,0,0))
```

# Finding the trend

## 1. Determine if there is actually any trend

(see Department of Water 2015).

### 1a. Apply the Mann-Kendall test from the 'trend' package

```{r Mann-Kendall test}
mk.test(coredata(soiltemp_T15_zoo))
#    or
# SeasonalMannKendall(soiltemp_T15_zoo) # needs base R regular time series object
```

### 1b. Estimate the Sen's slope

```{r Sens slope test}
sens.slope(coredata(soiltemp_T15_zoo))
```

## 2. Visualising a trend using a moving average

We create a new time series from a moving average for each 24h to remove daily
periodicity using the
[rollmean]{style="font-family: monospace; background-color: #e8e8e8;"} function
from the [zoo]{style="font-family: monospace; background-color: #e8e8e8;"}
package. We know these are hourly data with diurnal fluctuation, so a rolling
mean for chunks of length 24 should be OK, and this seems to be true based on 
Figure \@ref(fig:plot-movAv). In some cases the
[findfrequency()]{style="font-family: monospace; background-color: #e8e8e8;"}
function from the
[xts]{style="font-family: monospace; background-color: #e8e8e8;"} package can
detect the periodic frequency for us. A moving average will always smooth our
data, so a smooth curve doesn't necessarily mean that we have periodicity
(seasonality).

```{r moving average}
ff <- findfrequency(soiltemp_T15_xts) # often an approximation!! check the data!
cat("Estimated frequency is",ff,"\n")
```

```{r plot-movAv, out.width="55%", fig.align='center', fig.cap="Plots of soil temperature time series data and the 24-hour moving average.", results='hold'}
soiltemp_T15_movAv <- rollmean(soiltemp_T15_zoo, 24)
plot(soiltemp_T15_zoo, col=8, type="l") # original data
# add the moving average
lines(soiltemp_T15_movAv, lwd = 2)
legend("topright", legend = c("Data","Moving average"), col = c(8,1), lwd = c(1,2))
```

## 3. Showing a trend using a linear (regression) model

First we create a linear model of the time series...

```{r make linear model}
lm0 <- lm(coredata(soiltemp_T15_zoo) ~ index(soiltemp_T15_zoo))
summary(lm0)
```

...then use a plot (Figure \@ref(fig:linear-trend))to look at the linear 
relationship.

```{r linear-trend, out.width="55%", fig.align='center', fig.cap="Plots of soil temperature time series data and the overall trend shown by a linear model.", results='hold'}
soiltemp_T15_lmfit <- zoo(lm0$fitted.values, index(soiltemp_T15_zoo))
plot(soiltemp_T15_zoo, col = 8, type = "l")
lines(soiltemp_T15_lmfit, col = 2, lty = 2)
legend("topright", legend = c("Data","Linear trend"), col = c(8,2), lty = c(1,2))
```

Figure \@ref(fig:linear-trend) shows that a linear model does not really capture 
the trend (ignoring the periodicity) very convincingly. The next option 'loess' 
smoothing option below tries to present the trend more accurately.

## 4. Using Locally Estimated Scatterplot Smoothing (loess)

(**loess**, sometimes called 'lowess' is a form of locally weighted
non-parametric regression to fit smooth curves to data). The amount of smoothing 
is controlled by the [span =]{style="font-family: monospace; background-color: #e8e8e8;"} parameter in the [loess.smooth()]{style="font-family: monospace; background-color: #e8e8e8;"} function (from base **R**).

**NOTE**: loess smoothing does not have the relationship to ARIMA that moving averaging does, but does allow us to separate periodicity from random error in time series decomposition.

```{r loess-trend, out.width="55%", fig.align='center', fig.cap="Plots of soil temperature time series data and the overall trend shown by a loess smoothing model.", results='hold'}
y_trend <- loess.smooth(index(soiltemp_T15_zoo), 
                           coredata(soiltemp_T15_zoo), 
                           span = 0.15, evaluation = length(soiltemp_T15_zoo))
plot(soiltemp_T15_zoo, col = 8, type = "l")
soiltemp_T15_trend <- zoo(y_trend$y, index(soiltemp_T15_zoo))
lines(soiltemp_T15_lmfit, col = 2, lty = 2)
lines(soiltemp_T15_trend, col = "skyblue", lwd = 3)
legend("topright", bty = "n", inset = 0.02, cex = 1.25, 
       legend = c("actual data","linear model","loess smoothed"), 
       col = c(8,2,"skyblue"), lty=c(1,2,1), lwd = c(1,1,3))
```

# Isolating the Time Series Periodicity

**NOTE THAT TIME SERIES DON'T ALWAYS HAVE PERIODICITY !**

To model the periodicity we need to understand the autocorrelation structure of
the time series. We can do this graphically, first by plotting the
autocorrelation function
[acf()]{style="font-family: monospace; background-color: #e8e8e8;"} which 
outputs a plot by default:

```{r acf-plot, fig.height=3.5, fig.width=6, out.width='65%', fig.align='center', fig.cap="The autocorrelation function plot for soil temperature time series data (the horizontal axis is in units of seconds).", results='hold'}
acf(soiltemp_T15_xts, main = "")
```

What does Figure \@ref(fig:acf-plot) tell you about autocorrelation in this time 
series?

Then plot the partial autocorrelation function
([pacf()]{style="font-family: monospace; background-color: #e8e8e8;"})

```{r pacf-plot, fig.height=3.5, fig.width=6, out.width='65%', fig.align='center', fig.cap="The partial autocorrelation function plot for soil temperature time series data (horizontal axis units are seconds).", results='hold'}
pacf(soiltemp_T15_xts, main = "")
```

Interpreting partial autocorrelations (Figure \@ref(fig:pacf-plot)) is more
complicated - refer to Long & Teetor (2019, *Section 14.15*). Simplistically,
partial autocorrelation allows us to identify which and how many
autocorrelations will be needed to model the time series data.

## We use the 'Box-Pierce test' for autocorrelation

The null hypothesis H<sub>0</sub> is that no autocorrelation exists at any lag distance (so 
p $\le$ 0.05 'rejects' H<sub>0</sub>):

```{r box tests}
Box.test(soiltemp_T15_xts)
Box.test(diff(soiltemp_T15_xts,1)) # 1 difference
Box.test(diff(diff(soiltemp_T15_xts,1),1)) # 2 differences
```

## Make a time series of the loess residuals

Remember that we made a loess model of the time series ... the residuals 
(Figure \@ref(fig:loess-resids)) can give us the combination of periodicity
component plus any random variation.

```{r loess-resids, out.width='55%', fig.align='center', fig.cap="An incomplete decomposition of the soil temperature time series into smoothed trend and periodicity plus error (noise) components.", results='hold'}
soiltemp_T15_periodic <- soiltemp_T15_zoo - soiltemp_T15_trend
plot(soiltemp_T15_trend, ylim = c(-2,20), lty = 3)
lines(soiltemp_T15_periodic, col = "coral", lwd = 2) # just to check
legend("topright", legend = c("LOESS trend", "Periodicity plus noise"), 
       col = c(1,"coral"), lwd = c(1,2), lty = c(3,1))
```

We can also use *less smoothing* in the loess function to *retain* periodicity; we 
adjust the [span&nbsp;=]{style="font-family: monospace; background-color: #e8e8e8;"} 
option (lower values of [span]{style="font-family: monospace; background-color: #e8e8e8;"} give less smoothing; we just need to experiment with different
values). The difference between the data and the less-smoothed loess should be
just 'noise' or '**error**'.

We first generate a loess model which is stored in the object 
[temp_LOESS2]{style="font-family: monospace; background-color: #e8e8e8;"}:

```{r less-smoothed loess model, results='hold'}
temp_LOESS2 <- loess.smooth(index(soiltemp_T15_zoo), 
                           coredata(soiltemp_T15_zoo), 
                           span = 0.012, evaluation = length(soiltemp_T15_zoo))
```

We then use the new loess model to make a time series which contains both
periodic and trend information:

```{r less-smoothed loess time series, results='hold'}
soiltemp_T15_LOESS2 <- zoo(temp_LOESS2$y, index(soiltemp_T15_zoo))
```

The difference between the data and the less-smoothed loess should be just
'noise' or 'error', so we make a new time series based on this difference:

```{r error time series, results='hold'}
soiltemp_T15_err <- soiltemp_T15_zoo - soiltemp_T15_LOESS2
```

## plot, setting y axis limits to similar scale to original data:
```{r loess-periodic, out.width='55%', fig.align='center', fig.cap="Incomplete decomposition of soil temperature time series data showing the smoothed *trend*, combined *trend + periodicity*, and *error* components.", results='hold'}
par(mar = c(4,4,1,1), mgp = c(1.7,0.3,0), tcl = 0.3, font.lab = 2)
plot(soiltemp_T15_trend, ylim = c(-2,20), lty = 2, lwd = 2, # from above
     xlab = "2014 Date", ylab = "Soil temperature (\u00B0C)")
lines(soiltemp_T15_LOESS2, col = 3, lwd = 2)
lines(soiltemp_T15_err, col = 6) # from a couple of lines above
legend("left", 
       legend = c("Trend (coarse LOESS)", 
                  "Periodicity + trend (fine LOESS)",
                  "Unexplained variation"), 
       col = c(1,3,6), lwd = c(2,2,1), lty = c(2,1,1), bty="n")
```

The periodicity by itself should be represented by the difference between the
very smoothed (trend) and less smoothed (trend + periodicity) loess, as
suggested by the two loess curves in Figure \@ref(fig:loess-periodic). We now
make yet another time series object to hold this isolated periodicity:

```{r make periodic ts}
soiltemp_T15_periodic2 <- soiltemp_T15_LOESS2 - soiltemp_T15_trend
```

## Plot everything to show the components of time series decomposition

```{r decomp-plots, fig.height=9, fig.width=8, fig.align='center', fig.cap="Time series decomposition into the three main components (periodicity, trend, and error) for soil temperature (\u00B0C) at 15 cm depth."}
plot(cbind(soiltemp_T15_zoo,soiltemp_T15_periodic2,
     soiltemp_T15_trend, soiltemp_T15_err), main = "", xlab = "Date in 2014", 
     cex.main = 1.5, yax.flip = TRUE, col = c(8,6,4,2), ylim = c(-2,22))
x1 <- (0.5*(par("usr")[2]-par("usr")[1]))+par("usr")[1]
y <- (c(0.2, 0.3, 0.6, 0.82)*(par("usr")[4]-par("usr")[3]))+par("usr")[3]
text(rep(x1,4), y, 
     labels = c("Unaccounted variation","Trend","Apparent periodicity","Data"), 
     col = c(2,4,6,8))
```

<p>&nbsp;</p>

Figure \@ref(fig:decomp-plots) shows the original time series data, and the 
three important components of the decomposed time series: the *periodicity*, the
*trend*, and the *error* or unexplained variation. We should remember that:

- this time series was relatively easy to decompose into its components, but not all time series will be so 'well-behaved';
- there are different ways in which we can describe the trend and periodicity (we have just used the loess smoothing functions for clear visualization).

[This ends our exploratory data analysis of time series, which leads us into
ARIMA forecast modelling.]{style="font-family: serif; background-color: #ffff00;"}

# Modelling time series with ARIMA

All the analysis of our data before ARIMA is really exploratory data analysis of
time series:

-   Does our time series have periodicity?
-   Can we get stationarity with a moving average?
-   Does our time series have autocorrelation?
-   Can we get stationarity by differencing?

All of these operations are possible components of **ARIMA** models!

[**We recommend using the [xts]{.underline} format of a time series in ARIMA
model functions and forecasting**.]{style="color: #8000e0;"}

## Use the [forecast]{style="font-family: monospace; background-color: #e8e8e8;"} R package to run an ARIMA model

```{r auto arima}
auto.arima(soiltemp_T15_xts, max.p = 3, max.q = 3, max.d = 0, 
           seasonal = TRUE)
```

The [auto.arima]{style="font-family: monospace; background-color: #e8e8e8;"}
function runs a complex algorithm (Hyndman *et al*. 2020) to automatically
select the best ARIMA model on the basis of the **Aikake Information Criterion**
(AIC), a statistic which estimates how much information is 'lost' in the model
compared with reality. The AIC combines how well the model describes the data
with a 'penalty' for increased complexity of the model. Using AIC, a better
fitting model might not be selected if it has too many predictors. The best
ARIMA models will have the **lowest** AIC (or AICc) value.

## Use output of [auto.arima()]{style="font-family: monospace; background-color: #e8e8e8;"} to run arima

The [auto-arima]{style="font-family: monospace; background-color: #e8e8e8;"} 
algorithm is not perfect! -- but it does provide a starting point for examining
ARIMA models.

The output of the [auto-arima]{style="font-family: monospace; background-color: #e8e8e8;"} 
function includes a description of the best model the algorithm
found, shown as [ARIMA(p,d,q)]{style="font-family: monospace; background-color: #e8e8e8;"}. The p,d,q refer to

<table border="0" cellpadding="2" cellspacing="2" style="width: 85%;">
  <tr style="background-color: #e0e0e0;">
    <td><strong>Parameter</strong></td>
    <td><strong>Meaning</strong></td>
    <td><strong>Informed by</strong></td>
  </tr>
  <tr>
    <td>p</td>
    <td>The number of autoregressive predictors</td>
    <td>Partial autocorrelation</td>
  </tr>
  <tr style="background-color: #e0e0e0;">
    <td>d</td>
    <td>The number of differencing steps</td>
    <td>Stationarity tests &plusmn; differencing</td>
  </tr>
  <tr>
    <td>q</td>
    <td>The number of moving average predictors</td>
    <td>Stationarity tests &plusmn; moving averages</td>
  </tr>
</table>

&nbsp;<br>
A periodic or *seasonal* ARIMA model (often called SARIMA) has a more complex
specification: <br>
[ARIMA(p,d,q)(P,D,Q)(n)]{style="font-family: monospace; background-color: #e8e8e8;"} , <br>
where the additional parameters refer to the seasonality: **P** is the number of
seasonal autoregressive predictors, **D** the seasonal differencing, **Q** the
seasonal moving averages, and **n** is the number of time steps per
period/season.

### 1. With no seasonality

```{r simple arima, results='hold'}
am0 <- arima(x = soiltemp_T15_xts, order = c(1,0,3))
summary(am0)
confint(am0)
```

The output from [summary(am0)]{style="font-family: monospace; background-color: #e8e8e8;"} shows the values and uncertainties of the model parameters ([ar1,
ma1, ma2, ma3, intercept]{style="font-family: monospace;"}), the goodness-of-fit
parameters (we're interested in [aic]{style="font-family: monospace;"}). 

*Note*: [ar]{style="font-family: monospace;"} parameters are auto-regression
coefficients, and [ma]{style="font-family: monospace;"} are moving average
coefficients.

The output from [confint(am0)]{style="font-family: monospace; background-color: #e8e8e8;"} is a table showing the 95% confidence interval for the parameters. 
The confidence intervals should not include zero! (if so, this would mean we
can't be sure if the parameter is useful or not).

### 2. With seasonality

```{r seasonal arima, results='hold'}
ff <- findfrequency(soiltemp_T15_xts)
cat("Estimated time series frequency is",ff,"\n")
am1 <- arima(x = soiltemp_T15_xts, order = c(1,0,2),
             seasonal = list(order = c(1, 0, 1), period = ff))
summary(am1)
confint(am1)
```

*Note*: [sar]{style="font-family: monospace;"} parameters are *seasonal*
auto-regression coefficients, and [sma]{style="font-family: monospace;"} are
*seasonal* moving average coefficients.

The model which includes periodicity has the lowest AIC value (which is not
surprising, since the soil temperature data have clear periodicity).

Checking residuals using the [checkresiduals()]{style="font-family: monospace; background-color: #e8e8e8;"} function from the [forecast]{style="font-family: monospace; background-color: #e8e8e8;"} package is our best diagnostic tool to
assess the validity of our models. [This is a separate issue from how well the
model describes the data, measured by AIC.]

In the output (plot and text) from [checkresiduals()]{style="font-family: monospace; background-color: #e8e8e8;"}:
- the residual plot (top) should look like white noise
- the residuals should not be autocorrelated (bottom left plot)
- the p-value from the Ljung-Box test should be > 0.05 (text output)
- the residuals should be normally distributed (bottom right plot)

```{r diagnostic-arima, out.width="70%", fig.cap="Residual diagnostic plots for a non-seasonal ARIMA model of the  soil temperature time series.", results='hold'}
checkresiduals(am0)
```

```{r diagnostic-sarima, out.width="70%", fig.cap="Residual diagnostic plots for a seasonal ARIMA (SARIMA) model of the  soil temperature time series.", results='hold'}
checkresiduals(am1)
```

In both the diagnostic plots (Figure \@ref(fig:diagnostic-arima) and Figure 
\@ref(fig:diagnostic-sarima)) the residuals appear to be random and normally 
distributed. There seems to be a more obvious autocorrelation of residuals for 
the non-seasonal model, reinforcing the conclusion from the AIC value that the 
**seasonal** ARIMA model is the most appropriate.

## Use the ARIMA model to produce a forecast using both models

The whole point of generating ARIMA models is so that we can attempt to forecast
the future trajectory of our time series based on our existing time series data.

To do this, we use the appropriately named [forecast()]{style="font-family: monospace; background-color: #e8e8e8;"} function from the 
[forecast]{style="font-family: monospace; background-color: #e8e8e8;"} package.
The option [h =]{style="font-family: monospace; background-color: #e8e8e8;"} is 
to specify how long we want to forecast for after the end of our data -- in this 
case 168 hours (same units as the periodicity from [findfrequency]{style="font-family: monospace; background-color: #e8e8e8;"}), that is, one week.

```{r forecast objects}
fc0 <- forecast(am0, h = 168)
fc1 <- forecast(am1, h = 168)
```

## Then, look at forecasts with plots

```{r forecast-plots, fig.height=6, fig.width=7, out.width="70%", fig.align='center', fig.cap="Time series forecasts for soil temperature, based on (a) a non-seasonal ARIMA model and (b) a seasonal ARIMA model.", results='hold'}
par(mfrow = c(2, 1), cex.main = 0.9, mar = c(0,3,0,1), oma = c(3,0,1,0),
    mgp=c(1.6,0.3,0), tcl=0.2, font.lab=2)
plot(fc0,ylab = "Temperature (\u00B0C)",
     fcol = 4, xlab="", main = "", xaxt="n")
lines(soiltemp_T15_zoo, col = "grey70")
mtext("ARIMA with no seasonality", 1, -2.2, adj = 0.1, font = 2)
mtext(am0$call, 1, -1, adj = 0.1, col = 4, cex = 0.9, 
      family="mono", font = 2)
mtext("(a)", 3, -1.2, cex = 1.2, font = 2, adj = 0.95)

plot(fc1,ylab = "Temperature (\u00B0C)", main = "", fcol = 6)
lines(soiltemp$conc, col = "grey70")
mtext("Time since start (seconds)", 1, 1.7, cex = 1.2, font = 2)
mtext(paste("ARIMA with",am1$arma[5],"h periodicity"),
      side = 1, line = -2.2, adj = 0.1, font = 2)
mtext(am1$call, 1, -1, adj = 0.1, col = 6, cex = 0.9, family="mono")
mtext("(b)", 3, -1.2, cex = 1.2, font = 2, adj = 0.95)
```

From the two plots in Figure \@ref(fig:forecast-plots), we observe that the
seasonal ARIMA model seems to reflect the form of the original data much better,
by simulating the diurnal fluctuations in soil temperature. Although the
forecasting uncertainty shown by the shaded area in Figure
\@ref(fig:forecast-plots) is large for both types of ARIMA model, the interval
seems smaller for the seasonal ARIMA model (Figure
\@ref(fig:forecast-plots)(b)).

```{r reset mfrow 2, message=FALSE, warning=FALSE, include=FALSE}
par(mfrow = c(1, 1))
```

We can also make a slightly 'prettier' time series forecast plot using the 
[autoplot()]{style="font-family: monospace; background-color: #e8e8e8;"} 
function from the [ggplot2]{style="font-family: monospace; background-color: #e8e8e8;"} package (Figure \@ref(fig:forecast-ggplot)):

```{r forecast-ggplot, fig.height=4, fig.width=6, out.width='70%', fig.align='center', fig.cap="Time series forecast for soil temperature, based on  a seasonal ARIMA model, plotted using the ggplot2 autoplot() function.", results='hold'}
require(ggplot2) # gives best results using autoplot(...)
autoplot(fc1)+
  ylab("Temperature (\u00B0C)") +
  xlab(paste("Time since",index(soiltemp_T15_zoo)[1],"(s)")) +
  ggtitle("") +
  theme_bw()
```

[ARIMA models are not the end of the time series modelling and forecasting story!]{style="font-family: serif; font-size: 14pt; background-color: #ffff00;"}

# Exponential Smoothing Models

Sometimes, ARIMA models may not be the best option, and another commonly used
method is **exponential smoothing**.

## Try an exponential smoothing model

Check [help(forecast::ets)]{style="font-family: monospace; background-color: #e8e8e8;"} 
to correctly specify the model type using the
[model =]{style="font-family: monospace; background-color: #e8e8e8;"} option!

The easiest option (and the one used here) is to specify [model = "ZZZ"]{style="font-family: monospace; background-color: #e8e8e8;"}, which 
chooses the type of model automatically.

```{r make ets model}
soiltemp_T15_ets <- ets(soiltemp_T15_xts, model = "ZZZ")
summary(soiltemp_T15_ets)
```

In this case the [ets()]{style="font-family: monospace; background-color: #e8e8e8;"} function has automatically selected a model 
([ETS(A,Ad,N)]{style="font-family: monospace; background-color: #e8e8e8;"}) with
additive trend and errors, but no seasonality. We don't do it here, but we could
manually select a model with seasonality included.

The default plot function in R can plot forecasts from exponential smoothing
models too (Figure \@ref(fig:ets-forecast)).

```{r ets-forecast, out.width="70%", fig.align='center', fig.cap="Forecast of soil temperature based on an automatically selected exponential smoothing model without a seasonal component."}
plot(forecast(soiltemp_T15_ets, h=168), col=6, 
     xlab = "Time since start (s)", ylab = "Temperature (\u00B0C)", main="")
mtext(names(soiltemp_T15_ets$par),3,seq(-1,-5,-1), adj = 0.7, col = 4, font = 3)
mtext(signif(soiltemp_T15_ets$par,3),3,seq(-1,-5,-1), adj = 0.8, col = 4, font = 3)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
par(mfrow = c(1,1), mar = c(4,4,1,1), oma = c(0,0,0,0))
```

Exponential smoothing also [decomposes]{.underline} time series in a different
way, into *level* and *slope* components.

```{r ets-decomposition, fig.height=6, fig.width=6, out.width='75%', fig.align='center', fig.cap="Exponential smoothing time series decomposition plots for soil temperature data."}
plot(soiltemp_T15_ets, col = 2, xlab = "Time since start (s)") # ETS decomposition plots
```

<p style="text-align: center;">[(*Exponential smoothing is maybe not so good for the soil temperature data!*)]{style="font-family: serif; font-size: 14pt"}</p>

# References

Department of Water (2015). *Calculating trends in nutrient data*, Government of
Western Australia, Perth.
<https://water.wa.gov.au/__data/assets/pdf_file/0014/6800/Calculating-trends-in-nutrient-data-10-3-15.pdf>

Hyndman R, Athanasopoulos G, Bergmeir C, Caceres G, Chhay L, O'Hara-Wild M,
Petropoulos F, Razbash S, Wang E, Yasmeen F (2020). *forecast: Forecasting*
*functions for time series and linear models*. R package version 8.12,
<http://pkg.robjhyndman.com/forecast>.

Long, J.D., Teetor, P., 2019. Time series analysis. Chapter 14, *The R Cookbook*,
Second Edition <https://rc2e.com/timeseriesanalysis>.

R Core Team (2019). *R: A language and environment for statistical computing*. 
R Foundation for Statistical Computing, Vienna, Austria. URL
<https://www.R-project.org/>.

Ryan, J.A. and Ulrich, J.M. (2018). *xts: eXtensible Time Series*. R package
version 0.11-2. <https://CRAN.R-project.org/package=xts>

Zeileis, A. and Grothendieck, G. (2005). zoo: S3 Infrastructure for Regular and
Irregular Time Series. *Journal of Statistical Software*, **14**(6), 1-27.
<https://doi.org/10.18637/jss.v014.i06>
