---
title: "ENVT5503 Time Series"
subtitle: "Learning time series analysis concepts using hourly soil temperature data"
documentclass: article
classoption: a4paper
geometry: margin=0.787in
output: 
  pdf_document: 
  highlight: espresso
  fig_height: "50%"
  fig_width: "50%"
  fig_caption: TRUE
  number_sections: TRUE
  toc: TRUE
---

# Set up the R environment for time series analysis

We'll need a few packages for specialized time series and other functions in R...

```{r setup run, message=FALSE, warning=FALSE, include=TRUE, results='hide'}
# Load the packages we need 
library(zoo)      # for basic irregular time series functions
library(xts)      # we need the xts "eXtended Time Series" format for some functions
library(Kendall)  # for trend analysis with Mann-Kendall test
library(trend)    # for trend analysis using the Sen slope
library(forecast) # for time series forecasting with ARIMA and exponential smoothing 
library(tseries)  # for assessing stationarity using Augmented Dickey-Fuller test
library(lmtest)   # for Breusch-Pagan heteroscedasticity test etc.
library(car)      # for various commonly-used functions
library(ggplot2)  # alternative to base R plots

# (optional) make a better colour palette than the R default!
palette(c("black","red3","green3","blue2",
          "darkcyan","purple","sienna","gray50"))
par(mfrow = c(3,1), mar = c(4,4,1,1), mgp = c(1.7,0.3,0), tcl = 0.3, font.lab=2)
```

```{r setup not run, message=FALSE, warning=FALSE, include=FALSE}
# Load the packages we need for specialized time series and
# other functions in R
#
library(zoo)      # for basic irregular time series functions
library(xts)      # we need the xts "eXtended Time Series" format for some functions
library(Kendall)  # for trend analysis with Mann-Kendall test
library(trend)    # for trend analysis using the Sen slope
library(forecast) # for time series forecasting with ARIMA and exponential smoothing 
library(tseries)  # for assessing stationarity using Augmented Dickey-Fuller test
library(lmtest)   # for Breusch-Pagan heteroscedasticity test etc.
library(car)      # for various commonly-used functions
library(ggplot2)  # alternative to base R plots

# (optional) make a better colour palette than the R default!
palette(c("black","red3","green3","blue2",
          "darkcyan","purple","sienna","gray67"))
```

# Data input
Read the data into a data frame - this is how R often stores data - it's not the format we need but we'll use it for comparison.

## Non-time series object for comparison
```{r read into dataframe}
soiltemp <- read.csv("https://raw.githubusercontent.com/Ratey-AtUWA/Environmental-time-series/main/soiltemp2.csv")
colnames(soiltemp) <- c("Date","temp")
```

## Do some checks of the data
```{r check dataframe}
summary(soiltemp) # simple summary of each column
str(soiltemp) # more detailed information about the R object ('str'=structure)
```

### Check with a plot
```{r plot dataframe}
plot(soiltemp$temp, type = "l", col = 6)
```

We really want the data in a different type of R object! We use the '\texttt{zoo}' package to read the data in as a time series. This gives a data object of class 'zoo' which is much more flexible than the default time series object in R.

```{r read and summary csv to zoo}
soiltemp_T15_zoo <- read.csv.zoo("https://raw.githubusercontent.com/Ratey-AtUWA/Environmental-time-series/main/soiltemp2.csv",
                               format = "%Y-%m-%d %H:%M:%S", 
                               tz = "Australia/Perth", 
                               index.column=1,
                               header = TRUE)
# do some quick checks of the new R object:
summary(soiltemp_T15_zoo) 
str(soiltemp_T15_zoo) # POSIXct in the output is a date-time format
```

Check with a plot
```{r check plot zoo}
plot(soiltemp_T15_zoo, col = 2) 
```

Sometimes we need to use another time series data format, \textbf{xts} (e\underline{X}tended \underline{T}ime \underline{S}eries) which allows us to use more functions...

### make an xts object from our zoo object and check:

```{r make xts from zoo and check}
soiltemp_T15_xts <- as.xts(soiltemp_T15_zoo)
str(soiltemp_T15_xts) # just to check
plot(soiltemp_T15_xts, col = 6) # just to check
```

# Exploratory data analysis of time series

Try a plot of the data in our time series object, both the raw data and using common transformations. 
We compare with transformed time series data - we may need to do this to meet later modelling assumptions.

First we change the default plotting parameters using \texttt{par(...)}, then plot the object and some transformed versions (with custom axis labels).

```{r compare transformations plot, fig.height=7, fig.width=7}
par(mfrow = c(3,1), mar = c(4,4,1,1), mgp = c(1.7,0.3,0), tcl = 0.3, font.lab=2)

plot(soiltemp_T15_zoo, ylab = "Temperature (\u00B0C)",
     xlab = "Date", col = 7, lwd = 2, cex.lab = 1.4)

plot(log10(soiltemp_T15_zoo), 
     ylab = expression(bold(paste(log[10],"(Temperature, \u00B0C)"))),
     xlab = "Date", col = 6, lwd = 2, cex.lab = 1.4)

pt0 <- powerTransform(coredata(soiltemp_T15_zoo))
if(pt0$lambda<0) {
  plot(
    -1 * (soiltemp_T15_zoo ^ pt0$lambda),
    ylab = "power-transf. (Temp., \u00B0C)",
    xlab = "Date",
    col = 4, lwd = 2, cex.lab = 1.4
  )
} else {
  plot((soiltemp_T15_zoo ^ pt0$lambda),
       ylab = "power-transf. (Temp., \u00B0C)",
       xlab = "Date",
       col = 4, lwd = 2, cex.lab = 1.4
  )
}
rm(pt0)
```

Which of these plots looks like it might be homoscedastic, i.e., have constant variance regardless of time?

The R Cookbook suggests Box-Cox (power) transforming the variable to "stabilize the variance" (i.e. reduce heteroscedasticity). This does work, but can be better to use $log_{10}$[conc], as the values are easier to interpret.

If we think a transformation is needed, then run something like the code below (which does a square root transformation, i.e. $variable^{0.5}$):
```{r transform if needed not run, eval=FALSE, echo=TRUE}
soiltemp_T15_zoo <- soiltemp_T15_zoo^0.5 
# don't forget the xts version either!
soiltemp_T15_xts <- as.xts(soiltemp_T15_zoo)
```

# Assessing if a time series variable is stationary

A \textbf{stationary} variable's mean and variance are not dependent on time. In other words, for a stationary series, the mean and variance at any time are representative of the whole series.

If there is a trend for the value of the variable to increase or decrease, or if there are periodic fluctuations, we don't have a stationary time series.

Many useful statistical analyses and models for time series models need a stationary time series as input, or a time series that can be made stationary with transformations or differencing.

## Testing for stationarity

```{r adf tests}
# we need the package 'tseries' for the Augmented Dickey–Fuller (adf) Test
d0 <- adf.test(soiltemp_T15_zoo); print(d0)
d1 <- adf.test(diff(soiltemp_T15_zoo,1)); print(d1)
d2 <- adf.test(diff(diff(soiltemp_T15_zoo,1),1)); print(d2)
```
\newpage
## plot differencing for stationarity

```{r plot differenced series, fig.height=8, fig.width=8}
par(mfrow = c(3,1), mar = c(0,4,0,1), oma = c(4,0,1,0), cex.lab = 1.4)
plot(soiltemp_T15_zoo, ylab = "Raw data, no differencing",
     xlab="", xaxt="n", col = 8)
lines(loess.smooth(index(soiltemp_T15_zoo),coredata(soiltemp_T15_zoo)), 
      col = 4, lwd = 2)
abline(lm(coredata(soiltemp_T15_zoo)~index(soiltemp_T15_zoo)), col = 2)
legend("topright", legend = c("soiltemp_T15_Data", "Loess smoothing","Linear model"),
       cex = 1.8, col = c(1,4,2), lwd = c(1,2,1), bty = "n")
mtext(paste("adf.test p value =",signif(d0$p.value,3)), 
      side = 1, line = -1.2, adj = 0.05)
plot(diff(soiltemp_T15_zoo,1),
     ylab = "First differencing",
     xlab="", xaxt="n", col = 8)
abline(h = 0, col = "grey", lty = 2)
lines(loess.smooth(index(diff(soiltemp_T15_zoo,1)),coredata(diff(soiltemp_T15_zoo,1))), 
      col = 4, lwd = 2)
abline(lm(coredata(diff(soiltemp_T15_zoo,1))~index(diff(soiltemp_T15_zoo,1))), col = 2)
mtext(paste("adf.test p value =",signif(d1$p.value,3)), 
      side = 1, line = -1.2, adj = 0.05)
plot(diff(diff(soiltemp_T15_zoo,1),1),
     ylab = "Second differencing",
     xlab="Date", col = 8)
mtext("Date",side = 1, line = 2.2, font = 2)
abline(h = 0, col = "grey", lty = 2)
lines(loess.smooth(index(diff(diff(soiltemp_T15_zoo,1),1)),
                   coredata(diff(diff(soiltemp_T15_zoo,1),1))), 
      col = 4, lwd = 2)
abline(lm(coredata(diff(diff(soiltemp_T15_zoo,1)))~index(diff(diff(soiltemp_T15_zoo,1)))), col = 2)
mtext(paste("adf.test p value =",signif(d2$p.value,3)), 
      side = 1, line = -1.2, adj = 0.05)
```

```{r reset mfrow 1, message=FALSE, warning=FALSE, include=FALSE}
par(mfrow = c(1,1), mar = c(4,4,1,1), oma = c(0,0,0,0))
```

\newpage
# Finding the trend

## 1. Determine if there is actually any trend 
(see Department of Water 2015).

### 1a. apply the Mann-Kendall test from the 'trend' package
```{r Mann-Kendall test}
mk.test(coredata(soiltemp_T15_zoo))
#    or
# SeasonalMannKendall(soiltemp_T15_zoo) # needs base R regular time series object
```

### 1b. estimate the Sen's slope
```{r Sens slope test}
sens.slope(coredata(soiltemp_T15_zoo))
```

## 2. Visualising a trend using a moving average 
We create a new time series from a moving average for each 24h to remove daily periodicity using the \texttt{zoo::rollmean} function. We know these are hourly data with diurnal fluctuation, so a rolling mean for chunks of length 24 should be OK. In some cases the \texttt{findfrequency()} function from the xts package can detect the periodic frequency for us. A moving average will always smooth our data, so a smooth curve doesn't necessarily mean that we have periodicity (seasonality).

```{r moving average}
ff <- findfrequency(soiltemp_T15_xts) # often an approximation!! check the data!
cat("Estimated frequency is",ff,"\n")
soiltemp_T15_movAv <- rollmean(soiltemp_T15_zoo, 24)
plot(soiltemp_T15_zoo, col=8, type="l") # original data
# add the moving average
lines(soiltemp_T15_movAv, lwd = 2)
legend("topright", legend = c("Data","Moving average"), col = c(8,1), lwd = c(1,2))
```

## 3. Showing a trend using a linear (regression) model

First we create a linear model of the time series...
```{r make linear model}
lm0 <- lm(coredata(soiltemp_T15_zoo) ~ index(soiltemp_T15_zoo))
summary(lm0)
```

...then use a plot to look at the linear fit
```{r plot linear model}
soiltemp_T15_lmfit <- zoo(lm0$fitted.values, index(soiltemp_T15_zoo))
plot(soiltemp_T15_zoo, col = 8, type = "l")
lines(soiltemp_T15_lmfit, col = 2, lty = 2)
legend("topright", legend = c("Data","Linear trend"), col = c(8,2), lty = c(1,2))
```

## 4. using Locally Estimate Scatterplot Smoothing (loess)
(a form of locally weighted non-parametric regression to fit smooth curves to data)
```{r male and plot coarse loess}
y_trend <- loess.smooth(index(soiltemp_T15_zoo), 
                           coredata(soiltemp_T15_zoo), 
                           span = 0.15, evaluation = length(soiltemp_T15_zoo))

plot(soiltemp_T15_zoo, col = 8, type = "l")
soiltemp_T15_trend <- zoo(y_trend$y, index(soiltemp_T15_zoo))
lines(soiltemp_T15_lmfit, col = 2, lty = 2)
lines(soiltemp_T15_trend, col = "skyblue", lwd = 3)
legend("topright", bty = "n", inset = 0.02, cex = 1.25, 
       legend = c("actual data","linear model","loess smoothed"), 
       col = c(8,2,"skyblue"), lty=c(1,2,1), lwd = c(1,1,3))
```

# Isolating the Periodicity

\textbf{NOTE THAT TIME SERIES DON'T ALWAYS HAVE PERIODICITY !}

To model the periodicity we need to understand the autocorrelation structure of the time series. We can do this graphically, first by plotting the autocorrelation function \texttt{acf()}:
```{r acf plot, fig.height=3.5, fig.width=6, out.width='80%',out.height='80%'}
acf(soiltemp_T15_xts)
```

What does this tell you about autocorrelation in this time series?

Then plot the partial autocorrelation function (\texttt{pacf})
```{r pacf plot, fig.height=3.5, fig.width=6, out.width='80%',out.height='80%'}
pacf(soiltemp_T15_xts)
```

Interpreting partial autocorrelations is more complicated - refer to Long & Teetor (2019, Section 14.15). Simplistically, partial autocorrelation allows us to identify which and how many autocorrelations will be needed to model the time series data.

## We use the 'Box-Pierce test' for autocorrelation

The null hypothesis is that no autocorrelation exists at any lag distance (so p $\le$ 0.05 'rejects' null):

```{r box tests}
Box.test(soiltemp_T15_xts)
Box.test(diff(soiltemp_T15_xts,1)) # 1 difference
Box.test(diff(diff(soiltemp_T15_xts,1),1)) # 2 differences
```

... remember we made a loess model of the time series ... the residuals can give us the periodicity component plus random variation

## make a time series of the loess residuals
```{r loess resids}
soiltemp_T15_periodic <- soiltemp_T15_zoo - soiltemp_T15_trend
plot(soiltemp_T15_trend, ylim = c(-2,20), lty = 3)
lines(soiltemp_T15_periodic, col = "coral", lwd = 2) # just to check
legend("topright", legend = c("LOESS trend", "Periodicity plus noise"), 
       col = c(1,"coral"), lwd = c(1,2), lty = c(3,1))
```

We can use less smoothing in the loess function to retain periodicity; we adjust the 'span' option (lower values give less smoothing). The difference between the data and the less smoothed loess should be just 'noise' or 'error'.


```{r find periodic with 2x loess}
conc_LOESS2 <- loess.smooth(index(soiltemp_T15_zoo), 
                           coredata(soiltemp_T15_zoo), 
                           span = 0.012, evaluation = length(soiltemp_T15_zoo))
# then use the new loess model to make a time series...
# ...which contains both periodic and trend information
soiltemp_T15_LOESS2 <- zoo(conc_LOESS2$y, index(soiltemp_T15_zoo))

# the difference between the data and the less smoothed loess 
# should be just 'noise' or 'error'
soiltemp_T15_err <- soiltemp_T15_zoo - soiltemp_T15_LOESS2

# plot, setting y axis limits to similar scale to original data:
plot(soiltemp_T15_trend, ylim = c(-2,20), lty = 2, lwd = 2) # from above
lines(soiltemp_T15_LOESS2, col = 3, lwd = 2)
lines(soiltemp_T15_err, col = 6) # from a couple of lines above
legend("left", 
       legend = c("Trend (coarse LOESS)", 
                  " + Periodicity (fine LOESS)",
                  "Unexplained variation"), 
       col = c(1,3,6), lwd = c(2,2,1), lty = c(2,1,1), bty="n")
```

The periodicity should be represented by the difference between the very smoothed (trend) and less smoothed (trend + periodicity) loess.
```{r make periodic ts}
soiltemp_T15_periodic2 <- soiltemp_T15_LOESS2 - soiltemp_T15_trend
```

## Plot everything to show the components of time series decomposition
```{r decomposition plot, fig.height=9, fig.width=8}
plot(cbind(soiltemp_T15_zoo,soiltemp_T15_periodic2,
     soiltemp_T15_trend, soiltemp_T15_err),
     main = "Time series decomposition:\nSoil Temperature at 15cm (\u00B0C)", 
     cex.main = 1.5, yax.flip = TRUE, col = c(8,6,4,2), ylim = c(-2,22))
x1 <- (0.5*(par("usr")[2]-par("usr")[1]))+par("usr")[1]
y <- (c(0.2, 0.3, 0.6, 0.82)*(par("usr")[4]-par("usr")[3]))+par("usr")[3]
text(rep(x1,4), y, 
     labels = c("Unaccounted variation","Trend","Apparent periodicity","Data"), 
     col = c(2,4,6,8))
```

This ends our exploratory data analysis of time series, which leads us into ARIMA forecast modelling.

# Modelling time series with ARIMA

All the analysis of our data before ARIMA is really exploratory data analysis of time series 

\begin{itemize}
\item Does our time series have periodicity?
\item Can we get stationarity with a moving average?
\item Does our time series have autocorrelation?
\item Can we get stationarity by differencing?
\end{itemize}

All of these operations are possible components of \textbf{ARIMA} models!

\textbf{We recommend using the xts format of a time series in ARIMA model functions and forecasting}.

## Use the forecast:: R package to run an ARIMA model
```{r auto arima}
auto.arima(soiltemp_T15_xts, max.p = 3, max.q = 3, max.d = 0, 
           seasonal = TRUE)
```

The auto.arima function runs a complex algorithm (Hyndman *et al*. 2020) to automatically select the best ARIMA model on the basis of the \textbf{Aikake Information Criterion} (AIC), a statistic which combines how well the model describes the data with the degrees of freedom of the model. Using AIC, a better fitting model might not be selected if it has too many predictors. The best ARIMA models will have the LOWEST AIC (or AICc) value.

## Use output of auto.arima() to run arima

The auto-arima algorithm is not perfect! -- but it does provide a starting point for examining ARIMA models.

The output of the \texttt{auto-arima} function includes a description of the best model the algorithm found, shown as \\
\texttt{ARIMA(p,d,q)} \\
The p,d,q refer to 
\begin{tabular}{cll}
\textbf{Parameter} & Meaning & Informed by\\
\hline
p & The number of autoregressive predictors & Partial autocorrelation\\
d & The number of differencing steps & Stationarity tests \u00B1 differencing\\
q & The number of moving average predictors & Stationarity tests +/- moving averages\\
\end{tabular}

A periodic or seasonal ARIMA model (often called SARIMA) has a more complex specification: \newline
\texttt{ARIMA(p,d,q)(P,D,Q)(n)} \newline
Where the additional parameters refer to the seasonality: **P** is the number of seasonal autoregressive predictors, **D** the seasonal differencing, **Q** the seasonal moving averages, and **n** is the number of time steps per period/season. 

###   1. with no seasonality

```{r simple arima}
am0 <- arima(x = soiltemp_T15_xts, order = c(1,0,3))
summary(am0)
confint(am0)
```

###   2. with seasonality
```{r seasonal arima}
ff <- findfrequency(soiltemp_T15_xts)
cat("Estimated time series frequency is",ff,"\n")
am1 <- arima(x = soiltemp_T15_xts, order = c(1,0,2),
             seasonal = list(order = c(1, 0, 1), period = 24))
summary(am1)
confint(am1)
```

Checking residuals is our best diagnostic tool...

\begin{enumerate}
\item residual plot (top) should look like white noise
\item residuals should not be autocorrelated (bottom left plot)
\item p-value from Ljung-Box test should be > 0.05
\item residuals should be normally distributed (bottom right plot)
\end{enumerate}

```{r diagnostic plot simple arima}
checkresiduals(am0)
```

```{r diagnostic plot seasonal arima}
checkresiduals(am1)
```

## use the ARIMA model to produce a forecast using both models
```{r forecast objects}
fc0 <- forecast(am0, h = 168)
fc1 <- forecast(am1, h = 168)
```

## then, look at forecasts with plots
```{r forecast plots, fig.height=7, fig.width=7, results='hold'}
par(mfrow = c(2, 1), cex.main = 0.9, mar = c(3,3,1,1),
    mgp=c(1.6,0.3,0), tcl=0.2, font.lab=2)
plot(fc0,ylab = "Temperature (\u00B0C)",
     fcol = 4, xlab="Date/Time code", main = "")
lines(soiltemp_T15_zoo, col = "grey70")
mtext("ARIMA with no seasonality", 1, -2.2, adj = 0.1, font = 2)
mtext(am0$call, 1, -1, adj = 0.1, col = 4, cex = 0.9, 
      family="mono", font = 2)

plot(fc1,ylab = "Temperature (\u00B0C)",
     xlab="Date/Time code", main = "", fcol = 6)
lines(soiltemp$conc, col = "grey70")
mtext(paste("ARIMA with",am1$arma[5],"h periodicity"),
      side = 1, line = -2.2, adj = 0.1, font = 2)
mtext(am1$call, 1, -1, adj = 0.1, col = 6, cex = 0.9, family="mono")
```

```{r reset mfrow 2, message=FALSE, warning=FALSE, include=FALSE}
par(mfrow = c(1, 1))
```

### we can also make a slightly 'prettier' plot using ggplot2
```{r forecast ggplot, fig.height=4, fig.width=6, out.width='80%'}
require(ggplot2) # gives best results using autoplot(...)
autoplot(fc1)+
  ylab("Temperature (\u00B0C)") +
  xlab(paste("Time since",index(soiltemp_T15_zoo)[1],"(s)")) +
  ggtitle("Forecasted conc (\u00B5g/L)") +
  theme_bw()
```

\begin{tabular}{|c|}
\hline
\textbf{ARIMA models are not the end of the time series modelling and forecasting story.}\\
\hline
\end{tabular}

# Exponential Smoothing Models

Sometimes, ARIMA models may not be the best option, and another commonly used method is \textbf{exponential smoothing}.

## try an exponential smoothing model

Check \texttt{help(forecast::ets)} to correctly specify the model type! 

The parameters for lower & upper are c("alpha","beta","gamma","phi").

```{r make ets model}
soiltemp_T15_ets <- ets(soiltemp_T15_xts, model = "ZZZ")
summary(soiltemp_T15_ets)
```

The default plot function in R can plot forecasts from exponential smoothing models too...

```{r plot ets forecast}
plot(forecast(soiltemp_T15_ets, h=168), col = 6, 
     xlab = "Time since start", ylab = "Temperature (\u00B0C)")
mtext(names(soiltemp_T15_ets$par),3,seq(-1,-5,-1), adj = 0.7, col = 4, font = 3)
mtext(signif(soiltemp_T15_ets$par,3),3,seq(-1,-5,-1), adj = 0.8, col = 4, font = 3)
par(mfrow = c(1,1), mar = c(4,4,1,1), oma = c(0,0,0,0))
```

Exponential smoothing also decomposes time series in a different way:

```{r ets decomposition plots, fig.height=6, fig.width=6, out.height='75%', out.width='75%'}
plot(soiltemp_T15_ets, col = 2) # ETS decomposition plots
```

\textit{(Exponential smoothing is maybe not so good for the soil temperature data!)}

# REFERENCES

Department of Water (2015). Calculating trends in nutrient data, Government of Western Australia, Perth. https://water.wa.gov.au/__data/assets/pdf_file/0014/6800/Calculating-trends-in-nutrient-data-10-3-15.pdf

Hyndman R, Athanasopoulos G, Bergmeir C, Caceres G, Chhay L, O'Hara-Wild M, Petropoulos F, Razbash S, Wang E, Yasmeen F (2020). forecast: Forecasting functions for time series and linear models. R package version 8.12, http://pkg.robjhyndman.com/forecast.

Long, J.D., Teetor, P., 2019. Time series analysis. Chapter 14, The R Cookbook, Second Edition https://rc2e.com/timeseriesanalysis.

R Core Team (2019). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.

Ryan, J.A. and Ulrich, J.M. (2018). xts: eXtensible Time Series. R package version 0.11-2. https://CRAN.R-project.org/package=xts

Zeileis, A. and Grothendieck, G. (2005). zoo: S3 Infrastructure for Regular and Irregular Time Series. Journal of Statistical Software, 14(6), 1-27. https://doi.org/10.18637/jss.v014.i06
